Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2025-02-24 paper/main.tex
shortened and cleaned up discussion on LDA vs BERT + LDA vs Jaccard sim

\documentclass[10pt]{article}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{geometry}
\usepackage[sort&compress, numbers]{natbib}
\usepackage{pxfonts}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{lineno}

\usepackage{xcolor}

@@

% supplemental tables
\newcommand{\questions}{1}
\newcommand{\topics}{2}
\newcommand{\matchTab}{3}

% supplemental figures
\newcommand{\topicWordWeights}{1}
\newcommand{\topicWeights}{2}
\newcommand{\forcesCorrs}{3}
\newcommand{\bosCorrs}{4}
\newcommand{\jaccard}{5}
\newcommand{\ldaVsBERT}{6}
\newcommand{\individualKnowledgeMapsA}{7}
\newcommand{\individualKnowledgeMapsB}{8}
\newcommand{\individualKnowledgeMapsC}{9}
\newcommand{\individualLearningMapsA}{10}
\newcommand{\individualLearningMapsB}{11}

\newcommand{\U}{{\fontfamily{serif}\selectfont\ensuremath{\mathrm{U}}}}

% simple command for inline comments
\newcommand{\comment}[1]{}
% italicize section names in \nameref and place in \mbox to prevent bug when
% text is split across lines or pages
\NewCommandCopy{\oldnameref}{\nameref}
\renewcommand{\nameref}[1]{\mbox{\textit{\oldnameref{#1}}}}

@@

Our work makes several contributions to the study of how people acquire
conceptual knowledge. First, from a methodological standpoint, our modeling
framework provides a systematic means of mapping out and characterizing
knowledge in maps that have infinite (arbitrarily many) numbers of coordinates,
and of ``filling out'' those maps using relatively small numbers of 
multiple-choice quiz questions. Our experimental finding that we can use these 
maps to predict responses to held-out questions has several psychological 
implications as well. For example, concepts that are assigned to nearby 
coordinates by the text embedding model also appear to be ``known to a similar 
extent'' (as reflected by participants' responses to held-out questions;
Fig.~\ref{fig:predictions}). This suggests that participants also
\textit{conceptualize} similarly the content reflected by nearby embedding
coordinates. How participants' knowledge falls off with spatial distance is
captured by the knowledge maps we infer from their quiz responses
(e.g., Figs.~\ref{fig:smoothness},~\ref{fig:knowledge-maps}). In other words,
our study shows that knowledge about a given concept implies knowledge about
related concepts, and we also show how estimated knowledge falls off with
distance in text embedding space.

In our study, we characterize the ``coordinates'' of participants' knowledge
using a relatively simple ``bag of words'' text embedding model~\citep[LDA;
][]{BleiEtal03}. More sophisticated text embedding models, such as
transformer-based models~\citep{ViswEtal17, DevlEtal18, ChatGPT, TouvEtal23}
can learn complex grammatical and semantic relationships between words,
higher-order syntactic structures, stylistic features, and more. We considered
using transformer-based models in our study, but we found that the text
embeddings derived from these models were surprisingly uninformative with
respect to differentiating or otherwise characterizing the conceptual content
of the lectures and questions we used. We suspect that this reflects a broader
challenge in constructing models that are high-resolution within a given domain
(e.g., the domain of physics lectures and questions) \textit{and} sufficiently
broad so as to enable them to cover a wide range of domains. For example, we
found that the embeddings derived even from much larger and more modern models
like BERT~\citep{DevlEtal18}, GPT~\citep{ViswEtal17}, LLaMa~\citep{TouvEtal23},
and others that are trained on enormous text corpora, end up yielding poor
resolution within the content space spanned by individual course videos
(Supp.~Fig.~\ldaVsBERT). Whereas the LDA embeddings of the lectures and
questions are ``near'' each other (i.e., the convex hull enclosing the two
lectures' trajectories is highly overlapping with the convex hull enclosing the
questions' embeddings), the BERT embeddings of the lectures and questions are
instead largely distinct (top row of Supp.~Fig.~\ldaVsBERT). The LDA embeddings
of the questions for each lecture and the corresponding lecture's trajectory
are also similar. For example, as shown in Fig.~\ref{fig:sliding-windows}C, the
LDA embeddings for \textit{Four Fundamental Forces} questions (blue dots)
appear closer to the \textit{Four Fundamental Forces} lecture trajectory (blue
line), whereas the LDA embeddings for \textit{Birth of Stars} questions (green
dots) appear closer to the \textit{Birth of Stars} lecture trajectory (green
line). The BERT embeddings of the lectures and questions do not show this
property (Supp.~Fig.~\ldaVsBERT). We also examined per-question ``content
matches'' between individual questions and individual moments of each lecture
(Fig.~\ref{fig:question-correlations}, Supp.~Fig.~\ldaVsBERT). The time series
plot of individual questions' correlations are different from each other when
computed using LDA (e.g., the traces can be clearly visually separated), whereas
the correlations computed from BERT embeddings of different questions all look
very similar. This tells us that LDA is capturing some differences in content
between the questions, whereas BERT is not. The time series plots of individual
questions' correlations have clear ``peaks'' when computed using LDA, but not
when computed using BERT. This tells us that LDA is capturing a ``match''
between the content of each question and a relatively well-defined time window
of the corresponding lectures. The BERT embeddings appear to blur together the
content of the questions versus specific moments of each lecture. Finally, we
also compared the pairwise correlations between embeddings of questions within
versus across content areas (i.e., content covered by the individual lectures,
lecture-specific questions, and by the ``general physics knowledge''
questions). The LDA embeddings show a strong contrast between same-content
embeddings versus across-content embeddings. In other words, the embeddings of
questions about the \textit{Four Fundamental Forces} material are highly
correlated with the embeddings of the \textit{Four Fundamental Forces} lecture,
but not with the embeddings of \textit{Birth of Stars}, questions about
\textit{Birth of Stars}, or general physics knowledge questions. We see a
similar pattern with the LDA embeddings of the \textit{Birth of Stars}
questions (Fig.~\ref{fig:topics}, Supp.~Fig.~\topicWeights). In contrast, the
BERT embeddings are all highly correlated with each other (Supp.
Fig.~\ldaVsBERT). Taken together, these comparisons illustrate how LDA (trained
on the specific content in question) provides both coverage of the requisite
material and specificity at the level of the content covered by individual
questions. BERT, on the other hand, essentially assigns both lectures and all
of the questions (which are all broadly about ``physics'') into a tiny region
of its embedding space, thereby blurring out meaningful distinctions between
different specific concepts covered by the lectures and questions. We note that
these are not criticisms of BERT (or other large language models trained on
large and diverse corpora). Rather, our point is that simple fine-tuned models
trained on a relatively small but specialized corpus can outperform much more
complicated models trained on much larger corpora, when we are specifically
interested in capturing subtle conceptual differences at the level of a single
course lecture or question. Of course if our goal had been to find a model that
generalized to many different content areas, we would expect our approach to
perform comparatively poorly relative to BERT or other much larger models. We
suggest that bridging the tradeoff between high resolution within each content
area versus the ability to generalize to many different content areas will be
an important challenge for future work in this domain.

@@

At the opposite end of the spectrum from large language models, one could also
imagine \textit{simplifying} some aspects of our LDA-based approach by
computing simple word overlap metrics. For example, the Jaccard similarity
between text $A$ and $B$ is computed as the number of unique words in the
intersection of words from $A$ and $B$ divided by the number of unique words in
the union of words from $A$ and $B$. In a supplementary analysis
(Supp.~Fig.~\jaccard), we compared the LDA-based question-lecture matches we
reported in Figure~\ref{fig:question-correlations} with the Jaccard similarities
between each question and each sliding window of text from the corresponding
lecture. As shown in Supplementary Figure~\jaccard, this simple word-matching
approach does not appear to capture the same level of specificity as the
LDA-based approach. Whereas the LDA-based approach often yields a clear peak in
the time series of correlations between each question and the corresponding
lecture, the Jaccard similarity-based approach does not. Furthermore, these
LDA-based matches appear to capture conceptual overlaps between the questions
and lectures (Supp.~Tab.~\matchTab), whereas simple word matching does not. For
example, one of the example questions examined in Supplementary
Figure~\jaccard~asks ``Which of the following occurs as a cloud of atoms gets
more dense?'' The LDA-based matches identify lecture timepoints where the
relevant \textit{topics} are discussed (e.g., when words like ``cloud,''
``atom,'' ``dense,'' etc., are mentioned \textit{together}). The Jaccard
similarity-based matches, on the other hand, are strong when \textit{any} of
these words are mentioned, even if they do not occur together.

We view our approach as occupying a sort of ``sweet spot,'' between much larger
language models and simple word matching-based approaches, that enables us to
capture the relevant conceptual content of course materials at an appropriate
semantic scale. Our approach enables us to accurately and consistently identify
each question's content in a way that also matches up with what is presented in
the lectures. In turn, this enables us to construct accurate predictions about
participants' knowledge of the conceptual content tested by held-out questions
(Fig.~\ref{fig:predictions}).

>>>>

\documentclass[10pt]{article}
\usepackage{amsmath}
\usepackage[english]{babel}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{geometry}
\usepackage[sort&compress, numbers]{natbib}
\usepackage{pxfonts}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{lineno}
\usepackage{xcolor}

@@

% supplementary tables
\newcommand{\questions}{1}
\newcommand{\topics}{2}
\newcommand{\matchTab}{3}
% supplementary figures
\newcommand{\topicWordWeights}{1}
\newcommand{\topicWeights}{2}
\newcommand{\forcesCorrs}{3}
\newcommand{\bosCorrs}{4}
\newcommand{\jaccard}{5}
\newcommand{\ldaVsBERT}{6}
\newcommand{\individualKnowledgeMapsA}{7}
\newcommand{\individualKnowledgeMapsB}{8}
\newcommand{\individualKnowledgeMapsC}{9}
\newcommand{\individualLearningMapsA}{10}
\newcommand{\individualLearningMapsB}{11}
% supplementary discussion
\newcommand{\suppResults}{\textit{Supplementary results}}

% simple command for inline comments
\newcommand{\comment}[1]{}

% italicize section names in \nameref and place in \mbox to prevent bug when
% text is split across lines or pages
\NewCommandCopy{\oldnameref}{\nameref}
\renewcommand{\nameref}[1]{\mbox{\textit{\oldnameref{#1}}}}

@@

Our work makes several contributions to the study of how people acquire
conceptual knowledge. First, from a methodological standpoint, our modeling
framework provides a systematic means of mapping out and characterizing
knowledge in maps that have infinite (arbitrarily many) numbers of coordinates,
and of ``filling out'' those maps using relatively small numbers of 
multiple-choice quiz questions. Our experimental finding that we can use these 
maps to predict responses to held-out questions has several psychological 
implications as well. For example, concepts that are assigned to nearby 
coordinates by the text embedding model also appear to be ``known to a similar 
extent'' (as reflected by participants' responses to held-out questions;
Fig.~\ref{fig:predictions}). This suggests that participants also
\textit{conceptualize} similarly the content reflected by nearby embedding
coordinates. How participants' knowledge ``falls off'' with spatial distance is
captured by the knowledge maps we infer from their quiz responses
(e.g., Figs.~\ref{fig:smoothness},~\ref{fig:knowledge-maps}). In other words,
our study shows that knowledge about a given concept implies knowledge about
related concepts, and we also show how estimated knowledge falls off with
distance in text embedding space.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In our study, we characterize the ``coordinates'' of participants' knowledge
using a relatively simple ``bag of words'' text embedding model~\citep[LDA;
][]{BleiEtal03}. More sophisticated text embedding models, such as
transformer-based models~\citep{ViswEtal17, DevlEtal18, ChatGPT, TouvEtal23},
can leverage additional textual information such as complex grammatical and 
semantic relationships between words, higher-order syntactic structures, 
stylistic features, and more. We considered using transformer-based models in 
our study, but we found that the text embeddings derived from these models were 
surprisingly uninformative with respect to differentiating or otherwise 
characterizing the conceptual content of the lectures and questions we used 
(see \suppResults). We suspect that this reflects a broader challenge in 
constructing models that are both high-resolution within a given domain (e.g., 
the domain of physics lectures and questions) \textit{and} sufficiently broad 
as to enable them to cover a wide range of domains. Essentially, these 
``larger'' language models learn these more complex features of language through 
pre-training on enormous and diverse text corpora. But as a result, their 
embedding spaces also ``span'' an enormous and diverse range of conceptual 
content, sacrificing a degree of specificity in their capacities to distinguish 
subtle conceptual differences within a more narrow range of content. 
In comparing our LDA model (trained specifically on the lectures used in our 
study) to a larger transformer-based model (BERT), we found that LDA provides 
both coverage of the requisite material and specificity at the level of 
individual questions, while BERT essentially relegates the contents of both 
lectures and all quiz questions (which are all broadly about ``physics'') to a 
tiny region of its embedding space, thereby blurring out meaningful distinctions 
between different specific concepts covered by the lectures and questions 
(Supp.~Fig.~\ldaVsBERT). We note that these are not criticisms of BERT, nor of 
other large language models trained on large and diverse corpora. Rather, our 
point is that simpler models trained on relatively small but specialized 
corpora can outperform much more complex models trained on much larger corpora 
when we are specifically interested in capturing subtle conceptual differences 
at the level of a single course lecture or question. On the other hand, if our 
goal had been to choose a model that generalized to many different content 
domains, we would expect our LDA model to perform comparatively poorly to BERT 
or other much larger models. We suggest that bridging this tradeoff will be
an important challenge for future work.

At the opposite end of the spectrum from large language models, one could also
imagine using an even \textit{simpler} ``model'' than LDA that relates the 
contents of course lectures and quiz questions through explicit word-overlap 
metrics (rather than similarities in the latent topics they exhibit). In a 
supplementary analysis (Supp.~Fig.~\jaccard), we compared the LDA-based 
question-lecture matches shown in Figure~\ref{fig:question-correlations} with
analogous matches based on the Jaccard similarity between each question's text 
and each sliding window from the corresponding lecture's transcript. Similarly 
to the embeddings derived from BERT, we found that this approach also blurred 
meaningful distinctions between concepts presented in different parts of each 
lecture and tested by different quiz questions. In this case, this lack of 
specificity reflects that meaningful distinctions between concepts do not tend
to emerge at the level of individual words.

In this way, we view our approach as occupying a sort of ``sweet spot'' between
much larger language models and simpler word-matching-based approaches, in that 
it enables us to capture the relevant conceptual content of course materials at 
the appropriate semantic scale. Our approach enables us to accurately and 
consistently identify each question's content in a way that matches it with 
specific content from the lectures and distinguishes it from other questions 
about similar content. In turn, this enables us to construct accurate 
predictions about participants' knowledge of the conceptual content tested by 
individual quiz questions (Fig.~\ref{fig:predictions}).

@@
